---
title: "Pinus rubisco evolution"
author: "Arthur Leung"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
here::i_am("14. analyze.Rmd")

ggplot2::theme_set(cowplot::theme_cowplot(line_size = 0.25, font_size = 10) +
                     theme(axis.ticks.length = unit(-3, "pt"),
                           axis.ticks = element_line(linewidth = 0.25),
                           strip.background = element_blank(),
                           strip.placement = "outside",
                           axis.text.x.top = element_blank(),
                           axis.text.y.right = element_blank(),
                           axis.title.x.top = element_blank(),
                           axis.title.y.right = element_blank()))
```

# Selection analyses

## LRTs and model parameters from PAML site analyses. 

The M1a vs. M0 comparison was used to determine whether there were a class of codon sites with nearly neutral evolution (0 \< ω \< 1), where ω is the ratio of nonsynonymous to synonymous codon substitutions. The M3 vs. M0 tested for variable ω between sites. The M2a vs. M1a and M8a vs. M7 comparisons were used were used to determined if there were codon sites under positive selection (ω \> 1). The M2a vs. M1a test is more stringent than M8 vs. M7, so to prevent false positives we used only the results of M2a for downstream analyses; also this was the better fitting model based on AIC.

```{r paml}

mlc <- readr::read_csv(here::here("paml_site_mlc.csv"))
knitr::kable(mlc)
```

## Model parameters from CSUBST analysis. 

Branch pairs with at least one convergent substitution are shown.

```{r csubst}


ancestral_aa <- readr::read_csv(here::here("aa_at_sites_ancestral.csv"))

csubst_site <- readr::read_csv(here::here("site", "csubst_site_merged.csv")) %>%
  filter(OCNany2spe > 0.95) %>%
  distinct() %>%
  mutate(Site = codon_site_alignment) %>% # align with full cds position
  dplyr::select(!codon_site_alignment) %>%
  full_join(ancestral_aa) %>% # get root aa
  filter(aa_1 != AA_Ancestral & aa_2 != AA_Ancestral) %>% # no reversals
  mutate(substitution_1 = paste0(aa_1_anc, Site, aa_1),
         substitution_2 = paste0(aa_2_anc, Site, aa_2)) %>% 
  rename(aa_root = AA_Ancestral) %>%
  dplyr::select(Site, branch_id_1, branch_id_2, aa_root, substitution_1, substitution_2) %>%
  tidyr::pivot_longer(
    cols = ends_with(c("_1", "_2")),
    names_to = c(".value", "id"),
    names_pattern = "(.*)_(1|2)$"
  ) %>% # pivot longer
  dplyr::select(!id) %>%
  distinct() %>% # distinct
  group_by(Site, aa_root, substitution) %>%
  summarise(branch_ids = paste(branch_id %>% sort, collapse = ", "))

writexl::write_xlsx(csubst_site, here::here("convergent_subs.xlsx"))

```

## Positively selected, convergent sites

```{r omega}
ancestral_aa <- readr::read_csv(here::here("aa_at_sites_ancestral.csv")) %>%
  select(Site, AA_Ancestral) %>%
  mutate(Site = as.character(Site))
rst <- readr::read_csv(here::here("paml_site_omega.csv")) %>%
  filter(NSSitesModel == 8) %>%
  mutate(Site = as.character(Site)) %>%
  filter(PosteriorProb > 0.95)
convergent_sites <- readr::read_csv(here::here("site/csubst_site_merged.csv")) %>%
  filter(OCNany2spe > 0.95) %>%
  distinct() %>%
  mutate(Site = as.character(codon_site_alignment)) %>% # align with full cds position
  dplyr::select(!codon_site_alignment) %>%
  full_join(ancestral_aa) %>%
  filter(aa_1 != AA_Ancestral & aa_2 != AA_Ancestral) %>%
  dplyr::select(branch_id_1, branch_id_2, node1, node2, Site, OCNany2spe)

combined_sites <- inner_join(rst, convergent_sites) %>%
  filter(PosteriorProb > 0.95)

readr::write_csv(combined_sites, here::here("pos_selected_conv_sites.csv")) 

rst_labels <- full_join(rst, convergent_sites) %>%
  mutate(PosSelectedSite = ifelse(PosteriorProb > 0.95, Site, ""),
         ConvergentSite = ifelse(OCNany2spe > 0.95, "*", "")) %>%
  mutate(ConvergentSite = tidyr::replace_na(ConvergentSite, "")) %>%
  mutate(Label = PosSelectedSite) %>%
  mutate(Label = paste(PosSelectedSite, ConvergentSite, sep = "")) %>%
  distinct(Label, .keep_all = TRUE) %>%
  filter(stringr::str_detect(Label, stringr::regex("^[:digit:]")))

fig_omega <- readr::read_csv(here::here("paml_site_omega.csv")) %>%
  filter(NSSitesModel == 8) %>% 
  ggplot(aes(x = as.numeric(Site), y = w)) +
  geom_ribbon(aes(ymin = w-SE, ymax = w+SE), fill = "grey60") +
  geom_line(linewidth = 0.25) +
  geom_point(data = filter(rst, PosteriorProb > 0.95),
             shape = 21,
             color = "black",
             fill = "black",
             size = 1,
             alpha = 0.5,
             stroke = 0.25) +
  geom_text(data = rst_labels %>% filter(as.numeric(Site) < 150),
            aes(label = Label),
            size = 2.5,
            nudge_x = 0, nudge_y = 0.5) +
  geom_text(data = rst_labels %>% filter(Site %in% c(375, 418)),
            aes(label = Label),
            size = 2.5,
            nudge_x = 0, nudge_y = 0.5) +
  geom_text(data = rst_labels %>% filter(Site %in% c(255, 474)),
            aes(label = Label),
            size = 2.5,
            nudge_x = 18, nudge_y = 0.5) +
  geom_text(data = rst_labels %>% filter(Site %in% c(265)),
            aes(label = Label),
            size = 2.5,
            nudge_x = 24, nudge_y = 0.25) +
  geom_text(data = rst_labels %>% filter(Site %in% c(251)),
            aes(label = Label),
            size = 2.5,
            nudge_x = -18, nudge_y = 0.5) +
  geom_text(data = rst_labels %>% filter(Site %in% c(254, 439, 471)),
            aes(label = Label),
            size = 2.5,
            nudge_x = 0, nudge_y = 1.2) +
  geom_text(data = rst_labels %>% filter(Site %in% c(449)),
            aes(label = Label),
            size = 2.5,
            nudge_x = 6, nudge_y = 0.5) +
  cowplot::theme_cowplot() +
  xlab("Codon site") + 
  # scale_x_continuous(limits = c(25, 525), breaks = seq(from = 50, to = 500, by = 50), expand = c(0,0)) +
  ylab(expression(italic(d)[N]*"/"*italic(d)[S]))
ggsave(here::here("plots/s3_dnds_pine.svg"),fig_omega, width = 4, height = 3)

```


# Climate analyses

## Extract climate

```{r worldclim}

# Read data
gbif <- readr::read_csv(here::here("gbif_data", "clean.csv")) %>%
  dplyr::filter(!if_any(c(decimalLongitude, decimalLatitude), ~ is.na(.))) %>%
  distinct(decimalLongitude, decimalLatitude, .keep_all = TRUE)
taxon_keys <- readr::read_csv(here::here("gbif_data", "taxon_keys.csv")) %>%
  dplyr::select(speciesKey, verbatim_name) %>%
  distinct()

# Create SpatialPointsDataFrame
gbif_coords <- gbif %>%
  dplyr::select(decimalLongitude, decimalLatitude) %>%
  sp::SpatialPointsDataFrame(coords = ., data = gbif)

here::i_am("pine_rubisco/13. get_gbif_data.R")
# Get list of Bioclim files, specify files of interest using regex
bioclim_files <- list.files(here::here("download_gbif", "worldclim"), 
                              pattern = ".tif") 
# Iterate over the list of files and raster each, returning a list of raster objects 
bioclim_raster <-  purrr::map(here::here("download_gbif", "worldclim", bioclim_files),
                              ~raster::raster(.))
# Stack the raster objects
bioclim_stack <- raster::stack(bioclim_raster)

# Extract data from the raster stack
bioclim_extracted_raw <- raster::extract(bioclim_stack, gbif_coords, method = 'simple', sp = TRUE) %>%
  as.data.frame() %>%
  rename(
    MeanTempAnnual = wc2.1_10m_bio_1,
    MaxTempWarmestMonth = wc2.1_10m_bio_5,
    PrecAnnual = wc2.1_10m_bio_12,
    TempJan = wc2.1_10m_tavg_01,
    TempFeb = wc2.1_10m_tavg_02,
    TempMar = wc2.1_10m_tavg_03,
    TempApr = wc2.1_10m_tavg_04,
    TempMay = wc2.1_10m_tavg_05,
    TempJun = wc2.1_10m_tavg_06,
    TempJul = wc2.1_10m_tavg_07,
    TempAug = wc2.1_10m_tavg_08,
    TempSep = wc2.1_10m_tavg_09,
    TempOct = wc2.1_10m_tavg_10,
    TempNov = wc2.1_10m_tavg_11,
    TempDec = wc2.1_10m_tavg_12,
    PrecJan = wc2.1_10m_prec_01,
    PrecFeb = wc2.1_10m_prec_02,
    PrecMar = wc2.1_10m_prec_03,
    PrecApr = wc2.1_10m_prec_04,
    PrecMay = wc2.1_10m_prec_05,
    PrecJun = wc2.1_10m_prec_06,
    PrecJul = wc2.1_10m_prec_07,
    PrecAug = wc2.1_10m_prec_08,
    PrecSep = wc2.1_10m_prec_09,
    PrecOct = wc2.1_10m_prec_10,
    PrecNov = wc2.1_10m_prec_11,
    PrecDec = wc2.1_10m_prec_12
    ) %>%
  dplyr::select(!verbatim_name) %>% # species key = query term, from the name from the phylogeny > genbank accession
  right_join(., taxon_keys, by = "speciesKey", relationship = "many-to-many") %>%
  filter(!if_any(c(MeanTempAnnual, PrecAnnual), ~ is.na(.x))) %>%
  tidyr::pivot_longer(cols = starts_with("Temp") | starts_with("Prec") & !PrecAnnual,
                      names_to = c(".value", "Month"),
                      names_pattern = "(Temp|Prec)(.{3}$)")
bioclim_extracted <- bioclim_extracted_raw %>%
  filter(Temp >= 5) %>%
  # filter(Temp >= 5 & (Prec >= 2 * MeanTempAnnual)) %>%
  arrange(match(Month, month.abb)) %>%
  group_by(gbifID) %>%
  slice_max(order_by = Prec,
            n = 3,
            na_rm = TRUE) %>%
  mutate(GrowingSeasonTemp = mean(Temp), .keep = "unused") %>%
  ungroup()

here::i_am("13. get_gbif_data.R")

# Get short name (match phylogeny) and write file
bioclim <- bioclim_extracted %>%
  rename(Species = verbatim_name) %>%
  dplyr::select(c(Species, contains(c("Temp", "Prec")))) %>%
  mutate(ShortName = stringr::str_sub(stringr::word(Species, 2), 1, 10), .after = Species)
readr::write_csv(bioclim, here::here("gbif_data", "bioclim.csv"))

# Get summary statistics and write to file
bioclim_summary <- bioclim %>%
  group_by(Species, ShortName) %>%
  summarise(across(.cols = c(contains(c("Temp", "Prec"))), 
                   .fns = list(n = ~ length(!is.na(.)),
                               mean = ~ mean(., na.rm = TRUE),
                               se = ~ sd(., na.rm = TRUE)/length(!is.na(.)),
                               median = ~ median(., na.rm = TRUE),
                               q10 = ~ quantile(., 0.1, na.rm = TRUE),
                               q90 = ~ quantile(., 0.9, na.rm = TRUE)),
                   .names = "{.col}_{.fn}")) %>%
  filter(if_any(ends_with("_n"), ~ . >= 3))

bioclim_summary %>%
  ggplot(aes(x = MeanTempAnnual_mean, y = GrowingSeasonTemp_mean)) +
  geom_point() +
  scale_x_continuous(limits = c(0, NA)) +
  scale_y_continuous(limits = c(0, NA))

readr::write_csv(bioclim_summary, here::here("gbif_data", "bioclim_summary.csv"))
```

# Get PAML and amino acid data

```{r aa}

arc <- readr::read_csv(here::here("paml_site_arc.csv")) %>%
  filter(ID == 105) %>% # node 149 is the root node
  dplyr::select(Site, AA, PosteriorProb) %>%
  rename(AA_Ancestral = AA,
         AA_A_PosteriorProb = PosteriorProb)

pos_sel <- readr::read_csv(here::here("aa_at_sites.csv")) %>%
  tidyr::pivot_longer(cols = matches(stringr::regex("[:digit:]+")),
                      names_to = "Site",
                      values_to = "AA") %>%
  mutate(Site = as.numeric(Site),
         across(.cols = !Species,
                .fns = ~ifelse(. == "X", "-", .)))

aa <- inner_join(pos_sel, arc, by = "Site") %>%
  mutate(Ancestral = ifelse(AA == AA_Ancestral,
                            TRUE,
                            ifelse(AA == "-",
                                   NA,
                                   FALSE)))

sub_groups <- aa %>%
  filter(Ancestral == FALSE) %>%
  group_by(Site, AA_Ancestral, Ancestral) %>%
  summarise(Substitution = paste(unique(AA), collapse = "/")) %>%
  mutate(Substitution = paste(AA_Ancestral, Site, Substitution, sep = "")) %>%
  dplyr::select(!c(AA_Ancestral, Ancestral))

aa <- full_join(aa, sub_groups, by = c("Site", "AA_Ancestral"))

readr::write_csv(aa, here::here("aa_at_sites_ancestral.csv"))

```




## Leung et al.

### Read data

```{r clim_vars_stats}

bioclim_summary <- readr::read_csv(here::here("gbif_data", "bioclim_summary.csv")) %>%
  mutate(Species = ShortName)
climond_summary <- readr::read_csv(here::here("gbif_data", "climond_summary.csv")) %>%
  mutate(Species = ShortName)
aa <- readr::read_csv(here::here("aa_at_sites_ancestral.csv"))

sites_of_interest <- readr::read_csv(here::here("pos_selected_conv_sites.csv")) %>%
  pull(Site) %>%
  unique()

bioclim_aa <- inner_join(bioclim_summary, aa, by = "Species") %>%
  filter(!is.na(Ancestral),
         Site %in% sites_of_interest) %>%
  mutate(AA = reorder(AA, desc(Ancestral))) %>%
  dplyr::select(!contains(c("_median", "_se", "_q10", "_q90"))) %>%
  tidyr::pivot_longer(cols = contains("_mean"),
                      names_to = "ClimateVariable",
                      values_to = "ClimateValue") %>%
  group_by(ClimateVariable) %>%
  arrange(Site, Ancestral, ClimateValue, .by_group = TRUE) %>%
  distinct(Site, AA, ClimateVariable, ClimateValue, .keep_all = TRUE)
climond_aa <- inner_join(climond_summary, aa, by = "Species") %>%
  filter(!is.na(Ancestral),
         Site %in% sites_of_interest) %>%
  mutate(AA = reorder(AA, desc(Ancestral))) %>%
  dplyr::select(!contains(c("_median", "_se", "_q10", "_q90"))) %>%
  tidyr::pivot_longer(cols = contains("_mean"),
                      names_to = "ClimateVariable",
                      values_to = "ClimateValue") %>%
  group_by(ClimateVariable) %>%
  arrange(Site, Ancestral, ClimateValue, .by_group = TRUE) %>%
  distinct(Site, AA, ClimateVariable, ClimateValue, .keep_all = TRUE)

ancestral_aa <- bioclim_aa %>%
  ungroup() %>%
  select(Site, AA_Ancestral) %>%
  unique()

bioclim_aa_summary <- bioclim_aa %>%
  group_by(ClimateVariable, Site, AA) %>%
  summarise(across(.cols = ClimateValue,
                   .fns = list(n = length,
                               Mean = mean,
                               SE = ~sd(.)/length(.),
                               SD = ~sd(.)),
                   .names = "{.col}{.fn}"))
climond_aa_summary <- climond_aa %>%
  group_by(ClimateVariable, Site, AA) %>%
  summarise(across(.cols = ClimateValue,
                   .fns = list(n = length,
                               Mean = mean,
                               SE = ~sd(.)/length(.),
                               SD = ~sd(.)),
                   .names = "{.col}{.fn}"))
readr::write_csv(bind_rows(bioclim_aa_summary, climond_aa_summary), 
                 here::here("gbif_data", "bioclim_aa_summary.csv"))


# NONPHYLOGENETIC STATS

# pairwise_tests <- bioclim_aa %>%
#   group_by(ClimateVariable, Site) %>%
#   rstatix::t_test(formula = ClimateValue ~ AA, p.adjust.method =  "bonferroni", ref.group = "all") %>%
#   rstatix::add_significance(cutpoints = c(0, 0.001, 0.01, 0.05, 1),
#                             symbols = c("***", "**", "*", "")) %>%
#   # filter(p.adj.signif != "ns") %>%
#   group_by(ClimateVariable) %>%
#   rstatix::add_xy_position(fun = "mean_se", x = "AA", scales = "free", step.increase = 0) %>%
#   full_join(., ancestral_aa, by = "Site") %>%
#   mutate(AA = group1,
#          Ancestral = ifelse(AA == AA_Ancestral, TRUE, FALSE),
#          ClimateValue = y.position)
```

### Stats 

No need to run if already ran

```{r phy_anova1}
# PHYLOGENETIC ANOVA

tree <- ape::read.tree(here::here("Pinus_time_csubst.tre"))

stats <- bioclim_aa %>%
  filter(!(Site == 95 & AA == "A")) %>%
  select(ShortName, Site, AA, ClimateVariable, ClimateValue) %>%
  mutate(ClimateValue = sign(ClimateValue) * (abs(ClimateValue))^(1/3)) %>%
  group_by(Site, ClimateVariable) %>%
  tidyr::nest(.key = "Data") %>%
  mutate(Tree = purrr::map(Data, ~ ape::keep.tip(tree, .$ShortName))) %>%
  mutate(Groups = purrr::map(Data, ~ select(., "ShortName", "AA") %>% tibble::deframe())) %>%
  mutate(ResponseVar = purrr::map(Data, ~ select(., "ShortName", "ClimateValue") %>% tibble::deframe())) %>% # select + deframe is more readable than setNames
  rowwise() %>%
  # head(1) %>% # for testing with a smaller dataframe
  # mutate(ANOVAP = phytools::phylANOVA(tree = purrr::pluck(Tree),
  #                                    x = purrr::pluck(Groups) %>% droplevels(), # droplevels will drop the unused levels
  #                                    y = purrr::pluck(ResponseVar),
  #                                    posthoc = TRUE,
  #                                    nsim = 1000,
  #                                    p.adj = "none")$Pf) %>%  # Pt attribute has the adjusted P values
  mutate(PostHoc = phytools::phylANOVA(tree = purrr::pluck(Tree),
                                     x = purrr::pluck(Groups) %>% droplevels(), # droplevels will drop the unused levels
                                     y = purrr::pluck(ResponseVar),
                                     posthoc = TRUE,
                                     nsim = 1000,
                                     p.adj = "none")$Pt %>%  # Pt attribute has the adjusted P values
           as_tibble(rownames = "group1") %>%
           list()) %>% # list puts the tibble into an object of length 1 (to fit into one cell)
  mutate(PostHoc = tidyr::pivot_longer(PostHoc, cols = !contains("group"), names_to = "group2", values_to = "p") %>%
           list()) %>% # turn table with combos of groups and p values to group1, group2, p
  tidyr::unnest(PostHoc) %>%
  rowwise() %>%
  filter(group1 != group2) %>%
  mutate(groups = stringr::str_flatten(sort(c(group1, group2)), collapse = "-"), .before = "p") %>% 
  distinct(groups, .keep_all = TRUE) %>%
  group_by(Site, ClimateVariable) %>%
  mutate(p.adj = p.adjust(p, method = "fdr"))

climond_stats <- climond_aa %>%
  filter(!(Site == 95 & AA == "A")) %>%
  select(ShortName, Site, AA, ClimateVariable, ClimateValue) %>%
  mutate(ClimateValue = sign(ClimateValue) * (abs(ClimateValue))^(1/3)) %>%
  group_by(Site, ClimateVariable) %>%
  tidyr::nest(.key = "Data") %>%
  mutate(Tree = purrr::map(Data, ~ ape::keep.tip(tree, .$ShortName))) %>%
  mutate(Groups = purrr::map(Data, ~ select(., "ShortName", "AA") %>% tibble::deframe())) %>%
  mutate(ResponseVar = purrr::map(Data, ~ select(., "ShortName", "ClimateValue") %>% tibble::deframe())) %>% # select + deframe is more readable than setNames
  rowwise() %>%
  # head(1) %>% # for testing with a smaller dataframe
  # mutate(ANOVAP = phytools::phylANOVA(tree = purrr::pluck(Tree),
  #                                    x = purrr::pluck(Groups) %>% droplevels(), # droplevels will drop the unused levels
  #                                    y = purrr::pluck(ResponseVar),
  #                                    posthoc = TRUE,
  #                                    nsim = 1000,
  #                                    p.adj = "none")$Pf) %>%  # Pt attribute has the adjusted P values
  mutate(PostHoc = phytools::phylANOVA(tree = purrr::pluck(Tree),
                                     x = purrr::pluck(Groups) %>% droplevels(), # droplevels will drop the unused levels
                                     y = purrr::pluck(ResponseVar),
                                     posthoc = TRUE,
                                     nsim = 1000,
                                     p.adj = "none")$Pt %>%  # Pt attribute has the adjusted P values
           as_tibble(rownames = "group1") %>%
           list()) %>% # list puts the tibble into an object of length 1 (to fit into one cell)
  mutate(PostHoc = tidyr::pivot_longer(PostHoc, cols = !contains("group"), names_to = "group2", values_to = "p") %>%
           list()) %>% # turn table with combos of groups and p values to group1, group2, p
  tidyr::unnest(PostHoc) %>%
  rowwise() %>%
  filter(group1 != group2) %>%
  mutate(groups = stringr::str_flatten(sort(c(group1, group2)), collapse = "-"), .before = "p") %>% 
  distinct(groups, .keep_all = TRUE) %>%
  group_by(Site, ClimateVariable) %>%
  mutate(p.adj = p.adjust(p, method = "fdr"))

stats_adj <- stats %>%
  bind_rows(climond_stats) %>%
  select(Site, ClimateVariable, groups, p.adj) %>%
  group_by(Site, ClimateVariable) %>%
  tidyr::nest(.key = "PostHoc") %>%
  mutate(PostHoc = purrr::map(PostHoc, ~ tibble::deframe(.))) %>%
  filter(Site != 95) %>%
  mutate(Letters = purrr::map(PostHoc, ~ multcompView::multcompLetters(.)$Letters %>%
           as_tibble(rownames = "AA") %>%
           rename(Letter = value))) %>%
  tidyr::unnest(Letters) %>%
  rowwise() %>%
  mutate(Signif = any(purrr::pluck(PostHoc) < 0.05)) %>%
  mutate(Letter = ifelse(Signif, Letter, ""))

readr::write_csv(stats_adj, here::here("phy_anova.csv"))

```

# Pymol analyses

## Hydrogen bonds?

```{r h_bonds}
aa <- readr::read_csv(here::here("aa_at_sites_ancestral.csv")) %>% 
  filter(AA != "-") 
# Define residues that can form H-bonds, and also categorize amino acid types
H_BONDABLE <- c("S", # OH (hydroxyl)
                "T", # OH
                "Y", # OH
                "N", # CONH2 (amide)
                "Q", # CONH2
                "C", # SH (thiol) - not really, but can participate in dipole interactions with other thiols
                "H", # imidazole ring
                "D", # -COO- (carboxylate)
                "E", # -COO-
                "R", # guanidine
                "K") # -NH2 (amine)

ALIPHATIC <- c("A", "G", "I", "L", "P", "V")
AROMATIC <- c("F", "W", "Y")
ACIDIC <- c("D", "E")
BASIC <- c("R", "H", "K")
HYDROXYLIC <- c("S", "T")
SULFUR_CONTAINING <- c("C", "M")
AMIDIC <- c("N", "Q")
SITES_OF_INTEREST <- readr::read_csv(here::here("pos_selected_conv_sites.csv")) %>%
  pull(Site) %>%
  unique()

# See which residues go from non-H-bondable to H-bondable
h_bonds <- aa %>%
  filter(Site %in% SITES_OF_INTEREST, AA != "-") %>%
  ungroup() %>%
  select(Site, AA, AA_Ancestral) %>%
  unique() %>%
  filter(AA != AA_Ancestral) %>%
  mutate(AA_HBondable = ifelse(AA %in% H_BONDABLE, TRUE, FALSE),
         AA_A_HBondable = ifelse(AA_Ancestral %in% H_BONDABLE, TRUE, FALSE)) %>%
  rowwise() %>%
  mutate(across(c(AA, AA_Ancestral),
                .fns = ~ case_when(. %in% ALIPHATIC ~ "aliphatic",
                                   . %in% AROMATIC ~ "aromatic",
                                   . %in% ACIDIC ~ "acidic",
                                   . %in% BASIC ~ "basic",
                                   . %in% HYDROXYLIC ~ "hydroxylic",
                                   . %in% SULFUR_CONTAINING ~ "sulfur_containing",
                                   . %in% AMIDIC ~ "amidic"),
                .names = "{.col}_Type")) %>%
  filter(AA_HBondable == TRUE & AA_A_HBondable == FALSE |
           AA_HBondable == FALSE & AA_A_HBondable == TRUE) %>%
  arrange(AA_A_HBondable, AA_HBondable)
h_bonds

writexl::write_xlsx(h_bonds, here::here("h_bonds.xlsx"))

```

# Tables: Leaf habit and biome transitions ~ amino acids


``` {r}

library(phytools)


full_specific_epithet <- readr::read_csv(here::here("accessions.csv")) %>%
  dplyr::select(Species) %>%
  mutate(ShortName = stringr::str_sub(stringr::word(Species, 2), 1, 10), .before = "Species")

subs <- readr::read_csv(here::here("aa_at_sites_ancestral.csv")) %>%
  rowwise() %>%
  filter(!AA %in% c(AA_Ancestral, "-")) %>%
  distinct(Site, AA_Ancestral, Site, AA) %>%
  mutate(Substitution = paste0(AA_Ancestral, Site, AA)) %>%
  rename(AA_Derived = AA)

aa_root <- readr::read_csv(here::here("aa_at_sites_ancestral.csv")) %>%
  distinct(Site, AA_Ancestral) %>%
  mutate(Species = "ROOT",
         AA = AA_Ancestral,
         .keep = "unused") 
aa <- readr::read_csv(here::here("aa_at_sites_ancestral.csv")) %>%
  rename(ShortName = Species) %>% 
  full_join(full_specific_epithet, ., by = "ShortName") %>%
  filter(AA != "-") %>%
  ungroup() %>%
  select(Species, Site, AA) %>%
  bind_rows(aa_root) %>%
  full_join(subs) %>%
  mutate(Ancestral = ifelse(AA == AA_Derived, FALSE, TRUE))

tree <- ape::read.tree(here::here("Pinus_time_csubst.tre")) %>%
  treeio::rename_taxa(full_specific_epithet) %>%
  bind.tip(., tip.label = "ROOT", edge.length = 0, where = Ntip(.) + 1)
```

## Fire syndrome

```{r}

fire <- readxl::read_xlsx(here::here("jin_morphology.xlsx")) %>%
  mutate(Fire = ifelse(`Fire syndrome` == "Avoider", "a", "b")) %>%
  select(Species, Fire) %>%
  distinct(Species, .keep_all = TRUE) %>%
  add_row(Species = "ROOT", Fire = "a")

fire_lrts <- tibble(Substitution = character(),
                          model = character(),
                          Q_Ancestral = numeric(),
                          Q_Derived = numeric(),
                          deltaQ_Ancestral = numeric(),
                          deltaQ_Derived = numeric(),
                          FasterAA = character(),
                          StateFavouredbyAA = character(),
                          AIC_Dependent = numeric(),
                          AIC_Independent = numeric(),
                          deltaAIC = numeric(),
                          p = numeric())

for (SUB in subs %>% pull(Substitution)) {
  
  print(paste("fitPagels for substitution", SUB, "..."))
  
  aa_site <- aa %>%
    filter(Substitution == SUB) %>%
    mutate(Ancestral = ifelse(Ancestral, 0, 1)) %>%
    filter(Species != "ROOT") %>%
    select(Species, Ancestral)
  
  fire_species <- intersect(fire$Species, aa_site$Species) # get common species
  
  fire_fit_ER <- fitPagel(ape::keep.tip(tree, fire_species), # prune tree
                          aa_site %>% 
                            filter(Species %in% fire_species) %>% 
                            tibble::deframe(), # have to turn into named list
                          fire %>% 
                            filter(Species %in% fire_species) %>% 
                            tibble::deframe(),
                          dep.var = "y", 
                          model = "ER")
  fire_fit_SYM <- fitPagel(ape::keep.tip(tree, fire_species), # prune tree
                           aa_site %>% 
                             filter(Species %in% fire_species) %>% 
                             tibble::deframe(), # have to turn into named list
                           fire %>% 
                             filter(Species %in% fire_species) %>% 
                             tibble::deframe(),
                           dep.var = "y", 
                           model = "SYM")
  fire_fit_ARD <- fitPagel(ape::keep.tip(tree, fire_species), # prune tree
                           aa_site %>% 
                             filter(Species %in% fire_species) %>% 
                             tibble::deframe(), # have to turn into named list
                           fire %>% 
                             filter(Species %in% fire_species) %>% 
                             tibble::deframe(),
                           dep.var = "y", 
                           model = "ARD")
  
  fire_AICs <- tibble::tibble(model = "ER", 
                              AIC_dep = fire_fit_ER$dependent.AIC, 
                              AIC_indep = fire_fit_ER$independent.AIC) %>%
    add_row(model = "SYM", 
            AIC_dep = fire_fit_SYM$dependent.AIC, 
            AIC_indep = fire_fit_SYM$independent.AIC) %>%
    add_row(model = "ARD", 
            AIC_dep = fire_fit_ARD$dependent.AIC, 
            AIC_indep = fire_fit_ARD$independent.AIC)
  
  best_model <- fire_AICs %>%
    slice_min(AIC_dep, n = 1) %>%
    pull(model)
  
  if (best_model[1] == "ARD") {
    best_fire_fit <- fire_fit_ARD
  } else if (best_model[1] == "SYM") {
    best_fire_fit <- fire_fit_SYM
  } else {
    best_fire_fit <- fire_fit_ER
  }
  
  fire_lrts <- fire_lrts %>%
    add_row(Substitution = SUB,
            model = best_model[1],
            Q_Ancestral = best_fire_fit$dependent.Q["0|a", "0|b"],
            Q_Derived = best_fire_fit$dependent.Q["1|a", "1|b"],
            deltaQ_Ancestral = best_fire_fit$dependent.Q["0|a", "0|b"]-best_fire_fit$dependent.Q["0|b", "0|a"],
            deltaQ_Derived = best_fire_fit$dependent.Q["1|a", "1|b"]-best_fire_fit$dependent.Q["1|b", "1|a"],
            FasterAA = case_when(Q_Derived-Q_Ancestral > 0 ~ "derived",
                                      Q_Derived-Q_Ancestral < 0 ~ "ancestral",
                                      Q_Derived-Q_Ancestral == 0 ~ "same"),
            StateFavouredbyAA = case_when(deltaQ_Derived > 0 ~ "fire",
                                      deltaQ_Derived < 0 ~ "non-fire",
                                      deltaQ_Derived == 0 ~ "symmetrical"),
            AIC_Independent = best_fire_fit$independent.AIC[1],
            AIC_Dependent = best_fire_fit$dependent.AIC[1],
            deltaAIC = AIC_Independent-AIC_Dependent,
            p = best_fire_fit$P[1]) 
  
  writexl::write_xlsx(fire_lrts, here::here("pagels_test_fire.xlsx"))
  
}


```

## Fire syndrome, reverse

```{r}

fire <- readxl::read_xlsx(here::here("jin_morphology.xlsx")) %>%
  mutate(Fire = ifelse(`Fire syndrome` == "Avoider", "a", "b")) %>%
  select(Species, Fire) %>%
  distinct(Species, .keep_all = TRUE) %>%
  add_row(Species = "ROOT", Fire = "a")

fire_lrts <- tibble(Substitution = character(),
                          model = character(),
                          Q_Ancestral = numeric(),
                          Q_Derived = numeric(),
                          deltaQ_Ancestral = numeric(),
                          deltaQ_Derived = numeric(),
                          FasterAA = character(),
                          StateFavouredbyAA = character(),
                          AIC_Dependent = numeric(),
                          AIC_Independent = numeric(),
                          deltaAIC = numeric(),
                          p = numeric())

for (SUB in subs %>% pull(Substitution)) {
  
  print(paste("fitPagels for substitution", SUB, "..."))
  
  aa_site <- aa %>%
    filter(Substitution == SUB) %>%
    mutate(Ancestral = ifelse(Ancestral, 0, 1)) %>%
    filter(Species != "ROOT") %>%
    select(Species, Ancestral)
  
  fire_species <- intersect(fire$Species, aa_site$Species) # get common species
  
  fire_fit_ER <- fitPagel(ape::keep.tip(tree, fire_species), # prune tree
                          aa_site %>% 
                            filter(Species %in% fire_species) %>% 
                            tibble::deframe(), # have to turn into named list
                          fire %>% 
                            filter(Species %in% fire_species) %>% 
                            tibble::deframe(),
                          dep.var = "x", 
                          model = "ER")
  fire_fit_SYM <- fitPagel(ape::keep.tip(tree, fire_species), # prune tree
                           aa_site %>% 
                             filter(Species %in% fire_species) %>% 
                             tibble::deframe(), # have to turn into named list
                           fire %>% 
                             filter(Species %in% fire_species) %>% 
                             tibble::deframe(),
                           dep.var = "x", 
                           model = "SYM")
  fire_fit_ARD <- fitPagel(ape::keep.tip(tree, fire_species), # prune tree
                           aa_site %>% 
                             filter(Species %in% fire_species) %>% 
                             tibble::deframe(), # have to turn into named list
                           fire %>% 
                             filter(Species %in% fire_species) %>% 
                             tibble::deframe(),
                           dep.var = "x", 
                           model = "ARD")
  
  fire_AICs <- tibble::tibble(model = "ER", 
                              AIC_dep = fire_fit_ER$dependent.AIC, 
                              AIC_indep = fire_fit_ER$independent.AIC) %>%
    add_row(model = "SYM", 
            AIC_dep = fire_fit_SYM$dependent.AIC, 
            AIC_indep = fire_fit_SYM$independent.AIC) %>%
    add_row(model = "ARD", 
            AIC_dep = fire_fit_ARD$dependent.AIC, 
            AIC_indep = fire_fit_ARD$independent.AIC)
  
  best_model <- fire_AICs %>%
    slice_min(AIC_dep, n = 1) %>%
    pull(model)
  
  if (best_model[1] == "ARD") {
    best_fire_fit <- fire_fit_ARD
  } else if (best_model[1] == "SYM") {
    best_fire_fit <- fire_fit_SYM
  } else {
    best_fire_fit <- fire_fit_ER
  }
  
  fire_lrts <- fire_lrts %>%
    add_row(Substitution = SUB,
            model = best_model[1],
            Q_Ancestral = best_fire_fit$dependent.Q["0|a", "0|b"],
            Q_Derived = best_fire_fit$dependent.Q["1|a", "1|b"],
            deltaQ_Ancestral = best_fire_fit$dependent.Q["0|a", "0|b"]-best_fire_fit$dependent.Q["0|b", "0|a"],
            deltaQ_Derived = best_fire_fit$dependent.Q["1|a", "1|b"]-best_fire_fit$dependent.Q["1|b", "1|a"],
            FasterAA = case_when(Q_Derived-Q_Ancestral > 0 ~ "derived",
                                      Q_Derived-Q_Ancestral < 0 ~ "ancestral",
                                      Q_Derived-Q_Ancestral == 0 ~ "same"),
            StateFavouredbyAA = case_when(deltaQ_Derived > 0 ~ "fire",
                                      deltaQ_Derived < 0 ~ "non-fire",
                                      deltaQ_Derived == 0 ~ "symmetrical"),
            AIC_Independent = best_fire_fit$independent.AIC[1],
            AIC_Dependent = best_fire_fit$dependent.AIC[1],
            deltaAIC = AIC_Independent-AIC_Dependent,
            p = best_fire_fit$P[1]) 
  
  writexl::write_xlsx(fire_lrts, here::here("pagels_test_fire_reverse.xlsx"))
  
}

```


# Ancestral states

``` {r}

library(phytools)

full_specific_epithet <- readr::read_csv(here::here("accessions.csv")) %>%
  dplyr::select(Species) %>%
  mutate(ShortName = stringr::str_sub(stringr::word(Species, 2), 1, 10), .before = "Species")


dep_y_pagel <- readxl::read_xlsx(here::here("pagels_test_fire.xlsx")) %>%
  rowwise() %>%
  mutate(Min_AIC = min(AIC_Dependent, AIC_Independent)) %>%
  select(Substitution, Min_AIC, p)
dep_x_pagel <- readxl::read_xlsx(here::here("pagels_test_fire_reverse.xlsx"))  %>%
  rowwise() %>%
  mutate(Min_AIC_Re = min(AIC_Dependent, AIC_Independent)) %>%
  rename(p_Re = p) %>%
  select(Substitution, Min_AIC_Re, p_Re)
best_models <- full_join(dep_y_pagel, dep_x_pagel) %>%
  mutate(BestDepVar = case_when(Min_AIC < Min_AIC_Re & p < 0.05 ~ "y",
                                Min_AIC > Min_AIC_Re & p_Re < 0.05 ~ "x")) %>%
  filter(!is.na(BestDepVar)) %>%
  distinct(Substitution) %>%
  mutate(Site = stringr::str_extract(Substitution, "[:digit:]+") %>% as.numeric())
sites_of_interest <- best_models %>%
  pull(Site) %>%
  unique()
subs_of_interest <- best_models %>%
  pull(Substitution) %>%
  unique()

aa <- readr::read_csv(here::here("aa_at_sites_ancestral.csv")) %>%
  filter(Site %in% sites_of_interest) %>%
  rename(ShortName = Species) %>%
  full_join(full_specific_epithet, ., by = "ShortName") %>%
  filter(AA != "-") %>%
  select(Species, Site, AA) %>%
  full_join(best_models) %>%
  mutate(Ancestral = ifelse(AA == stringr::str_sub(Substitution, -1, -1), "1", "0")) %>%
  select(Species, Substitution, Ancestral)

tree <- ape::read.tree(here::here("Pinus_time_csubst.tre")) %>%
  treeio::rename_taxa(full_specific_epithet) %>%
  bind.tip(., tip.label = "ROOT", edge.length = 0, where = Ntip(.) + 1)

fire <- readxl::read_xlsx(here::here("jin_morphology.xlsx")) %>%
  mutate(Fire = ifelse(`Fire syndrome` == "Avoider", "0", "1")) %>%
  select(Species, Fire) %>%
  distinct(Species, .keep_all = TRUE) %>%
  add_row(Species = "ROOT", Fire = "0")

SUBSTITUION = "L375M"
OVERWRITE = FALSE

for (SUBSTITUTION in subs_of_interest) {

  file_list <- list.files(here::here("plots/simmaps"))
  
  if (OVERWRITE == FALSE & paste("fire", SUBSTITUTION, ".svg", sep = "") %in% file_list) {
    print("File exists, continuing")
    next
    }
  
  print(paste("Doing site", SUBSTITUTION))

  aa_site <- aa %>%
    filter(Substitution == SUBSTITUTION) %>%
    select(!Substitution)

  common_species <- intersect(pull(fire, Species), pull(aa_site, Species))

  pruned_tree <- tree %>% ape::keep.tip(common_species)

  ER_aa <- fitMk(pruned_tree,
                 aa_site %>% filter(Species %in% common_species) %>% tibble::deframe(),
                 model = "ER",
                 pi = c("0" = 1, "1" = 0))
  ARD_aa <- fitMk(pruned_tree,
                  aa_site %>% filter(Species %in% common_species) %>% tibble::deframe(),
                  model = "ARD",
                  pi = c("0" = 1, "1" = 0))
  if (AIC(ER_aa) < AIC(ARD_aa)) {
    print("AA: Choosing ER")
    sim_tree_aa <- simmap(ER_aa, nsim = 100)
  } else {
    print("AA: Choosing ARD")
    sim_tree_aa <- simmap(ARD_aa, nsim = 100)
  }
  dmap_aa <- densityMap(sim_tree_aa)
  dmap_aa_color <- setMap(dmap_aa, c("black", "#fde725"))

  ER_fire <- fitMk(pruned_tree,
                    fire %>% filter(Species %in% common_species) %>% tibble::deframe(),
                    model = "ER",
                    pi = c("0" = 1, "1" = 0))
  ARD_fire <- fitMk(pruned_tree,
                     fire %>% filter(Species %in% common_species) %>% tibble::deframe(),
                     model = "ARD",
                     pi = c("0" = 1, "1" = 0))
  if (AIC(ER_fire) < AIC(ARD_fire)) {
    print("fire: Choosing ER")
    sim_tree_fire <- simmap(ER_fire, nsim = 100)
  } else {
    print("fire: Choosing ARD")
    sim_tree_fire <- simmap(ARD_fire, nsim = 100)
  }
  dmap_fire <- densityMap(sim_tree_fire)
  dmap_fire_color <- setMap(dmap_fire, c("black", "#d42dfc"))

  svg(here::here("plots/simmaps", paste( "fire", SUBSTITUTION, ".svg", sep = "")),
      height = 5, width = 5)
  layout(matrix(1:2,1,2),widths=c(1,1))
  par(oma=c(0,0,0,0))
  par(mar=c(0,0,0,0))
  plot(dmap_aa_color, lwd = 2, ftype="off", legend = FALSE, no.margin=T)
  plot(dmap_fire_color, lwd = 2, direction="leftwards", ftype="off", legend = FALSE,no.margin=T)
  dev.off()
}


```

# Summary Table

```{r summarytable}
ancestral_aa <- readr::read_csv(here::here("aa_at_sites_ancestral.csv")) %>%
  select(Site, AA_Ancestral) %>%
  mutate(Site = as.character(Site))
rst <- readr::read_csv(here::here("paml_site_omega.csv")) %>%
  filter(NSSitesModel == 2) %>%
  mutate(Site = as.character(Site)) %>%
  filter(PosteriorProb > 0.95) %>%
  select(Site, PosteriorProb) %>%
  mutate(PosteriorProb = ifelse(PosteriorProb > 0.95, "*", ""))

convergent_sites <- readr::read_csv(here::here("site", "csubst_site_merged.csv")) %>%
  filter(OCNany2spe > 0.95) %>%
  distinct() %>%
  mutate(Site = as.character(codon_site_alignment)) %>% # align with full cds position
  dplyr::select(!codon_site_alignment) %>%
  full_join(ancestral_aa) %>%
  filter(aa_1 != AA_Ancestral & aa_2 != AA_Ancestral) %>%
  dplyr::select(branch_id_1, branch_id_2, node1, node2, Site, OCNany2spe) %>%
  distinct(Site, OCNany2spe) %>%
  group_by(Site) %>%
  slice_min(order_by = OCNany2spe) %>%
  mutate(OCNany2spe = ifelse(OCNany2spe > 0.95, "*", "")) %>%
  arrange(as.numeric(Site))

fire_lrts <- readxl::read_xlsx(here::here("pagels_test_fire.xlsx")) %>%
  arrange(Site)%>%
  mutate(Site = as.character(Site),
         Fire = case_when(p < 0.05 & StateFavoured == "non-fire" ~ "-",
                           p < 0.05 & StateFavoured == "fire" ~ "+",
                           p < 0.05 ~ "*",
                           .default = ""),
         .keep = "unused") %>%
  select(Site, Fire) 

h_bonds <- readxl::read_xlsx(here::here("h_bonds.xlsx")) %>%
  mutate(HydrogenBonds = ifelse(AA_HBondable & !AA_A_HBondable, "+", "-")) %>% # gaining h-bonds means more rigidity, thus better in warmer environments
  select(Site, HydrogenBonds) %>%
  mutate(Site = as.character(Site))

aovs_bisse <- readxl::read_xlsx(here::here("bisse.xlsx")) %>%
  group_by(Site) %>%
  mutate(Site = as.character(Site)) %>%
  slice_min(AIC, n = 1) %>%
  filter(Name != "equal sp. and ex. rate") %>%
  mutate(DiversificationRate = case_when(lambda1 - mu1 > lambda0 - mu0 ~ "+",
                                         lambda1 - mu1 < lambda0 - mu0 ~ "-",
                                         .default = "*")) %>%
  select(Site, DiversificationRate)

summary_table <- rst %>% rename(PositiveSelection = PosteriorProb) %>%
  full_join(convergent_sites %>% rename(ConvergentEvolution = OCNany2spe)) %>%
  full_join(fire_lrts) %>%
  full_join(h_bonds) %>%
  full_join(aovs_bisse) %>%
  relocate(HydrogenBonds, .after = "ConvergentEvolution") %>%
  mutate(across(.cols = everything(),
                .fns = ~ifelse(is.na(.), "", .))) %>%
  arrange(as.numeric(Site))

writexl::write_xlsx(summary_table, here::here("plots/1_summary_pine.xlsx"))

```

